---
title: "VOC"
author: "Anna Mrazova"
date: "2025-04-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Loading Data and libraries
## Formating
```{r Data and libraries}
library(readxl)
library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(glmmTMB)
library(FactoMineR)
library(factoextra)
library(MuMIn)
library(e1071)
library(lmtest)
library(DHARMa)
library(dplyr)
library(ggplot2)
library(vegan)
library(broom)
library(pheatmap)

setwd("C:/Users/annam/OneDrive - Biologické centrum AV ČR, v. v. i/Anawork/Students/Diana Tomandlova/Bakalarka/data")
comp_all <- read_excel("250404_Tomandlova_BP_VOC_Results.xlsx", sheet = "Results")

# Pivot to long format
comp_all_long <- comp_all %>%
  pivot_longer(
    cols = 5:ncol(.),
    names_to = "Compound",
    values_to = "Peak"
  ) %>%
  # Extract treatment (non-numeric) 
  mutate(treatment = str_extract(Desig, "^[^0-9]+"))

# Recode 'Bag' into actual numeric days
comp_all_long <- comp_all_long %>%
  mutate(
    Day = recode(as.character(Bag),
                 "1" = 0,
                 "2" = 15,
                 "3" = 30)
  )
```

## 1) Effect of treatment and day on overall VOC production
```{r Overall VOC production}

comp_all_long <- comp_all_long %>%
  group_by(treatment, Desig, Day) %>%
  summarise(total_peak = sum(Peak, na.rm = TRUE), .groups = "drop")

tr_bag <- lmer(total_peak ~ treatment * Day + (1|Desig), data = comp_all_long)
tr_bag_tmb <- glmmTMB(total_peak ~ treatment * Day + (1|Desig), data = comp_all_long, family = Gamma)

# We have too small sample size so the random effects make mess... Thus,
# we go for simple linear model:

full_model <- lm(total_peak ~ treatment * Day, data = comp_all_long)
```

### Diagnostics checks
```{r Diagnostic checks}
par(mfrow = c(2, 2))
plot(full_model)  # Residuals vs fitted, QQ-plot, etc.
skewness(residuals(full_model))  # Near 0 is best
shapiro.test(residuals(full_model))
bptest(full_model)  # Breusch-Pagan test
sim_res <- simulateResiduals(fittedModel = full_model, plot = TRUE)
```

#### Given the results, we have to log transform the data as normality: dispersion, residuals vs. fitted values and homoscedasticity is violated

### Data transformation and diagnostics
```{r Data log transform}
comp_all_long$total_peak_log <- log1p(comp_all_long$total_peak)

full_model_log <- lm(total_peak_log ~ treatment * Day, data = comp_all_long)

par(mfrow = c(2, 2))
plot(full_model_log) 
skewness(residuals(full_model_log))
shapiro.test(residuals(full_model_log))
bptest(full_model_log)
sim_res <- simulateResiduals(fittedModel = full_model_log, plot = TRUE)
```

#### Log transformation helped. We can continue with log-transformed data:

### Variable selection and model averaging
```{r}
options(na.action = "na.fail")
model_set_log <- dredge(full_model_log, subset = TRUE)
avg_model_log <- model.avg(model_set_log, subset = delta < 2)
summary(avg_model_log)
#Effect size:
exp(1.67)       # Effect of MeJA
exp(1.43)       # Effect of MeJA + Ptáci
```
#### We will interpret the data based on model averaging because:
#### **1)** There are two plausible models (ΔAICc < 2),
#### **2)** Our top model has less than ~70% weight,
#### **3)** We want to account for model uncertainty.

#### **Results:**
#### **1)** Day has a weak possitive effect on VOC production
#### **2)** MeJA treatment increases VOCs by 5.3 compared to Control (K)
#### **3)** MeJA + Ptáci also increases VOCs (by 4.2)

### Visualization 1
```{r Plot1}
# Summarize by treatment and day
summary_df <- comp_all_long %>%
  group_by(Day, treatment) %>%
  summarise(
    mean_peak = mean(total_peak, na.rm = TRUE),
    se_peak = sd(total_peak, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )
# Plot
voc_tr_day <- ggplot(summary_df, aes(x = Day, y = mean_peak, 
                                     color = treatment, fill = treatment)) +
  geom_jitter(data = comp_all_long, 
              aes(x = Day, y = total_peak, color = treatment),
              width = 1.5, height = 0, alpha = 0.4, size = 1.5, inherit.aes = FALSE) +
  geom_line(size = 1.2) +
  geom_ribbon(aes(ymin = mean_peak - se_peak, ymax = mean_peak + se_peak),
              alpha = 0.2, color = NA) +
  geom_point(size = 3, shape = 21, fill = "white") +
  labs(x = "Den", y = "Suma VOC",
       title = "Efekt ošetření a času na celkovou produkci VOC") +
  theme_minimal() +
  theme(
    axis.line = element_line(linewidth = 1, colour = "black"),
    axis.title.x = element_text(size = 20, face = "bold"),
    axis.title.y = element_text(size = 20, face = "bold"),
    axis.text = element_text(size = 18),
    plot.title = element_text(hjust = 0.5, size = 24, face = "bold")
  ) +
  scale_color_manual(values = c(
    "P" = "orange",
    "K" = "#2F80ED",
    "M+P" = "black",
    "M" = "red")
  ) +
  scale_fill_manual(values = c(
    "P" = "orange",
    "K" = "#2F80ED",
    "M+P" = "black",
    "M" = "red")
  )

print(voc_tr_day)
```

## 2) RD Analysis: differences in VOC profiles (qualitative)
### Data
```{r data}
comp_all <- read_excel("250404_Tomandlova_BP_VOC_Results.xlsx", sheet = "Results")

# Pivot back to wide
comp_wide <- comp_all %>%
  pivot_longer(cols = 5:ncol(.), names_to = "Compound", values_to = "Peak") %>%
  mutate(
    treatment = str_extract(Desig, "^[^0-9]+"),
    Day = recode(as.character(Bag), "1" = 0, "2" = 15, "3" = 30)
  ) %>%
  pivot_wider(names_from = Compound, values_from = Peak, values_fill = 0)

# VOCs only
voc_matrix <- comp_wide %>%
  select(-c(Desig, Bag, treatment, Day))

metadata <- comp_wide %>%
  select(Desig, treatment, Day) %>%
  mutate(treatment = factor(treatment),
         Day = as.numeric(Day))
```

### Helliger transformation and results
```{r Hell results}
voc_hell <- decostand(voc_matrix, method = "hellinger")

rda_model <- rda(voc_hell ~ treatment * Day, data = metadata)
summary(rda_model)
anova(rda_model, by = "terms", permutations = 999)
```
#### **Results:**
#### The whole model is significant -> VOC comopsition differs between treatments and/or time of sampling
#### The overall explained variance: ~40 % (constrained inertia = 0.102 / total = 0.2545)

#### Treatment: p < 0.001 -> has a significant effect on VOC composition
#### Day: p < 0.001 -> Sampling time also significantly contributes ti changes in VOC composition

### Visualization 2
```{r Plot2}
treatment_cols <- c(
  "K" = "#2F80ED",
  "M" = "red",
  "M+P" = "black",
  "P" = "orange"  
)

par(mfrow = c(1, 1))
plot(rda_model, type = "n", scaling = 2, main = "RDA – složení VOC")
points(rda_model, display = "sites",
       col = treatment_cols[metadata$treatment],
       pch = 19)
ordihull(rda_model, groups = metadata$treatment, draw = "polygon",
         col = treatment_cols, 
         border = treatment_cols, 
         lwd = 2, alpha = 80)
legend("topright", legend = names(treatment_cols), 
       col = treatment_cols, pch = 19, bty = "n", title = "Ošetření")
```

## 3) Effect of treatment and time on individual compounds
### Data
```{r}
comp_all <- read_excel("250404_Tomandlova_BP_VOC_Results.xlsx", sheet = "Results")

comp_long <- comp_all %>%
  pivot_longer(cols = 5:ncol(.), names_to = "Compound", values_to = "Peak") %>%
  mutate(
    treatment = str_extract(Desig, "^[^0-9]+"),
    Day = recode(as.character(Bag), "1" = 0, "2" = 15, "3" = 30),
    treatment = factor(treatment),
    Day = as.numeric(Day)
  )
```

### LM models for each compound
```{r LM models}

model_list <- comp_long %>%
  group_by(Compound) %>%
  group_map(~ lm(log1p(Peak) ~ treatment * Day, data = .x), .keep = TRUE)

names(model_list) <- unique(comp_long$Compound)
```

### Heatmap
```{r Heatmap}
model_results <- purrr::map_df(model_list, tidy, .id = "Compound") %>%
  filter(term %in% c("treatmentM", "treatmentP", "treatmentM+P", "Day",
                     "treatmentM:Day", "treatmentP:Day", "treatmentM+P:Day"))
estimates <- model_results %>%
  select(Compound, term, estimate) %>%
  pivot_wider(names_from = term, values_from = estimate)

pvals <- model_results %>%
  select(Compound, term, p.value) %>%
  pivot_wider(names_from = term, values_from = p.value)

annot <- as.matrix(pvals[,-1])
annot_stars <- ifelse(annot < 0.05, "*", "")
annot_combined <- matrix(paste0(round(annot, 3), annot_stars), 
                         nrow = nrow(annot), dimnames = dimnames(annot))

pheatmap(as.matrix(estimates[,-1]),
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         main = "Heatmapa účinků ošetření a času na jednotlivé sloučeniny (log)",
         display_numbers = annot_combined,
         color = colorRampPalette(c("blue", "white", "red"))(100),
         fontsize_number = 9,
         labels_row = estimates$Compound)
```

### **Results:**
#### For each of the 36 detected volatile organic compounds (VOCs), we fitted a linear model with the predictors treatment * Day using log-transformed response values (log1p(Peak)) to assess how individual compounds changed over time and in response to treatment.

#### The results were visualized using a heatmap showing the direction and magnitude of the estimated effects (red = positive effect, blue = negative effect), with p-values displayed as numerical annotations.

#### **1)** Strong and statistically significant effects were observed primarily for the MeJA and MeJA + ptáci treatments.

#### **2)** The most strongly affected compounds included Methyl salicylate, Unknown11, Unknown12, Alkane2, Caryophyllene, and Copaene.

#### **3)** The bird treatment alone (treatmentP) did not show any consistent effect on individual VOC production.

#### The table below summarizes the 25 most significantly affected compounds (p < 0.05), including the direction and magnitude of the treatment effects (on the log scale):

### Significantly different compounds per treatment
```{r Table results}
signif_effects <- c("treatmentM", "treatmentM+P")

significant_compounds_selected <- model_results %>%
  filter(term %in% signif_effects, p.value < 0.05) %>%
  select(Compound, term, estimate, p.value) %>%
  arrange(term, p.value)

print(significant_compounds_selected, n = 25)
```